---
title: "BIS15L_AI_hmwk"
author: "Your Name"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions
This homework is a bit less traditional this week. There are some errors in the code below. In addition add notes and titles to the code chunks that explains what the code chunck does. Use AI tools, notes from class, or any other resources to help fix the errors. Feel free to edit any plots to your preference. Don't forget to answer the two questions at the bottom. Upon completion, you should not have to '#' out any of the code chunks.

Push your final work to your repository. Your final lab report should be organized, clean, and run free from errors. Remember, you must remove the `#` for the included code chunks to run. Be sure to add your name to the author header above.   

Again make sure to use the formatting conventions of RMarkdown to make your report neat and clean!  


```{r}
library(tidyverse)
library(janitor)
library(dplyr)
```


1. Load the data
```{r}
blood_count <- read_csv("data/blood_count_dataset.csv") %>% clean_names()
```

2. Take a look at the data
```{r}
glimpse(blood_count)
```

3. What is the min, mean, and max of red blood cells by sex
```{r}
blood_count %>% 
  group_by(gender) %>% 
  summarize(min=min(red_blood_cells),
            mean=mean(red_blood_cells),
            max=max(red_blood_cells))
```


4. What is the mean of the blood count for males
```{r}
options(scipen = 999, digits = 2)

blood_count %>%
  filter(gender == "male") %>%
  select(3:9) %>%
  colMeans(na.rm = TRUE)
```

5. What is the mean of the blood count for females
```{r}
blood_count %>%
  filter(gender == "Female") %>%
  select(3:9) %>%
  colMeans(na.rm = TRUE)
```


6. Make a boxplot of the blood count tests
```{r}
blood_count %>%
  pivot_longer(cols = 3:9, names_to = "test", values_to = "values") %>%
  ggplot(aes(x = gender, y = values, fill = gender)) +
  geom_boxplot() +
  facet_wrap(~ test, scales = "free") +
  labs(title = "Boxplot of Blood Count Tests",
       x = "Test",
       y = "Values")
```

7. Group the data by age group
```{r}
grouped_bc <- blood_count %>%
  mutate(age_group = case_when(
    age >= 20 & age <= 29 ~ "20s",
    age >= 30 & age <= 39 ~ "30s",
    age >= 40 & age <= 49 ~ "40s",
    age >= 50 & age <= 59 ~ "50s",
    age >= 60 & age <= 70 ~ "60s")) %>% # there were only 4 people in 70s, so I combined it with 60s
  head()
grouped_bc
```

8. Make a boxplot of the blood count tests by age group
```{r}
grouped_bc %>%
  pivot_longer(cols = 3:9, names_to = "test", values_to = "values") %>%
  ggplot(aes(x = age_group, y = values, fill = age_group)) +
  geom_boxplot() +
  facet_wrap(~ test, scales = "free") +
  labs(title = "Boxplot of Blood Count Tests by Age Group",
       x = "Age Group",
       y = "Values")
```

9. What are two drawbacks and two benefits of AI?
**One drawback of AI is that it's not always correct, it makes mistakes and is biased. Another drawback is that it can cause some ethical concerns when used in certains feilds, like medicine. One benefit of AI is that it can help people when they are stuck on code or when they need help figuring out whats wrong with their code. Another benefit is that it can help speed up the process on projects when there is a bunch of repetitive code.**

10. Do you think you will use AI for coding in the future? Why or why not?
**Yes, I would use AI for coding in the future since it helps with complex coding problems and can help fix any mistakes in code.**



## Push your final code to GitHub!
Please be sure that you check the `keep md` file in the knit preferences. 