---
title: "lab 5 warmup"
output: html_document
date: "2024-01-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Warm-up
The warm-up will use the fish data from lab 4. These data are from: Gaeta J., G. Sass, S. Carpenter. 2012. Biocomplexity at North Temperate Lakes LTER: Coordinated Field Studies: Large Mouth Bass Growth 2006. Environmental Data Initiative.  [link](https://portal.edirepository.org/nis/mapbrowse?scope=knb-lter-ntl&identifier=267). 

You will need the tidyverse to complete the warm-up- be sure to load the library.  

```{r}
library("tidyverse")
```

### 1. Load the fish data. 

```{r}
getwd()
```

```{r}
fish <- read_csv("data/Gaeta_etal_CLC_data.csv")
```

```{r}
names(fish)
```

### 2. Transform the fish data to only include the variables lakeid and length.  Store this as a new dataframe called "fishlength".   

```{r}
fishlength <- select(fish,"lakeid", "length")
fishlength
```

### 3. Filter the `fish` data to include the samples from lake "BO".  
```{r}
filter(fish, lakeid == "BO")
```

### 4. Calculate the mean length of fish from lake "BO".  
```{r}
fish_BO <- filter(fish, lakeid == "BO") 
```

```{r}
mean(fish_BO$length)
```

