---
title: "Midterm 2 Master Notes"
author: "Lejla Becirevic"
date: "2024-02-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(skimr)
library(naniar)
library(visdat)
library(here)
library(palmerpenguins)

``` 

## LAB 8

**summarize() and group_by() are powerful tools that we can use to produce clean summaries of data. Especially when used together, we can quickly group variables of interest and save time**
```{r}
print(penguins)
```

**Produce a summary of the mean for bill_length_mm, bill_depth_mm, flipper_length_mm, and body_mass_g within Adelie penguins only. Be sure to provide the number of samples**
```{r}
penguins %>% 
  filter(species=="Adelie") %>% 
  summarize(mean_bill_length=mean(bill_length_mm, na.rm = T),
            mean_bill_depth=mean(bill_depth_mm, na.rm = T),
            mean_flippper_length=mean(flipper_length_mm, na.rm = T),
            mean_body_mass=mean(body_mass_g, na.rm = T),
            n=n())
```

### across()
**Function in dplyr called `across()` which can count for multiple variables**

**Use `summarize()` to produce distinct counts over multiple variables; i.e. species, island, and sex?** 
```{r}
penguins %>%
  summarize(distinct_species = n_distinct(species),
            distinct_island = n_distinct(island),
            distinct_sex = n_distinct(sex))
```

**By using `across()` we can reduce the clutter and make things cleaner**
```{r}
penguins %>%
  summarize(across(c(species, island, sex), n_distinct))
```

**This is very helpful for continuous variables**
```{r}
penguins %>%
  summarize(across(contains("mm"), mean, na.rm=T))
```

**group_by also works**
```{r}
penguins %>%
  group_by(sex) %>% 
  summarize(across(contains("mm"), mean, na.rm=T))
```

**Here we summarize across all variables**
```{r}
penguins %>%
  summarise_all(n_distinct)
```

**Operators can also work, here I am summarizing n_distinct() across all variables except species, island, and sex**
```{r}
penguins %>%
  summarize(across(!c(species, island, sex, year), 
                   mean, na.rm=T))
```

**All variables that include "bill"...all of the other dplyr operators also work**
```{r}
penguins %>%
  summarise(across(starts_with("bill"), n_distinct))
```

**Produce separate summaries of the mean and standard deviation for bill_length_mm, bill_depth_mm, and flipper_length_mm for each penguin species. Be sure to provide the number of samples**
```{r}
penguins %>% 
  group_by(species) %>% 
  summarise(across(c(contains("mm"), body_mass_g), mean, na.rm=T), #included variable after contains "mm" with the c() function, can also add stuff after
            n=n())
```

```{r}
getwd()
```

**Load the mammals life history data and clean the names**
```{r}
life_history <- read_csv("/Users/lejla.becirevic/Desktop/BIS15W2024_lbecirevic/homework/data/mammal_lifehistories_v3.csv") %>% clean_names()
```

**Here is a new one for you using the `purrr` package. This will give you a quick summary of the number of NA's in each variable.**
```{r}
life_history %>% 
  map_df(~ sum(is.na(.)))
```

**A single approach to deal with NA's in this data set**
```{r}
life_history <- read_csv("/Users/lejla.becirevic/Desktop/BIS15W2024_lbecirevic/homework/data/mammal_lifehistories_v3.csv", na= c("NA", " ", ".", "-999", "not measured")) %>% clean_names() # you nee to know how the NAs are represented in the data; you dont want to do this by default
```

### naniar
**naniar is a package that is built to manage NA's**

**miss_var_summary provides a clean summary of NA's across the data frame**
```{r}
naniar::miss_var_summary(life_history)
```

**Let's use mutate() and na_if() to replace 0's with NA's**
```{r}
life_history <- 
  life_history %>% 
  mutate(max_life=na_if(max_life, 0))
```

**We can also use `miss_var_summary` with `group_by()`. This helps us better evaluate where NA's are in the data.**
```{r}
life_history %>%
  group_by(order) %>%
  select(order, wean_mass) %>% 
  miss_var_summary(order=T)
```

**naniar also has a nice replace function which will allow you to precisely control which values you want replaced with NA's in each variable.**
```{r}
life_history %>% 
  replace_with_na(replace = list(newborn = "not measured", 
                                 weaning= -999, 
                                 wean_mass= -999, 
                                 afr= -999, 
                                 max_life= 0, 
                                 litter_size= -999, 
                                 gestation= -999, 
                                 mass= -999)) %>% 
miss_var_summary() #makes replacement of NAs specific with variable
```

**Import the data and do a little exploration. Be sure to clean the names if necessary.
```{r}
cites <- read_csv("/Users/lejla.becirevic/Desktop/BIS15W2024_lbecirevic/homework/data/cites.csv") %>% clean_names()
```

**Try using `group_by()` with `naniar`. Look specifically at class and `exporter_reported_quantity`. For which taxonomic classes do we have a high proportion of missing export data?**
```{r}
cites %>% 
  select(class, exporter_reported_quantity) %>% 
  group_by(class) %>%
  miss_var_summary() %>% 
  arrange(desc(pct_miss))
```

### Visualizing NAs
```{r}
vis_dat(life_history) #classes of data
```

```{r}
vis_miss(life_history)
```

### Dealing with NA's in advance
**If you are sure that you know how NA's are treated in the data, then you can deal with them in advance using `na()` as part of the `readr` package.**
```{r}
life_history_advance <- 
  readr::read_csv(file = "/Users/lejla.becirevic/Desktop/BIS15W2024_lbecirevic/homework/data/mammal_lifehistories_v3.csv", 
                  na = c("NA", " ", ".", "-999")) #all NA, blank spaces, .,and -999 are treated as NA
```


## LAB 9
### pivot_longer  
**Scientists frequently use spreadsheets that are organized to make data entry efficient. This is often referred to as wide format. Unfortunately, the wide format creates a problem because column names may actually represent values of a variable. The command `pivot_longer()` shifts data from wide to long format.**   

**Rules:** 
+ `pivot_longer`(cols, names_to, values_to)
+ `cols` - Columns to pivot to longer format
+ `names_to` - Name of the new column; it will contain the column names of gathered columns as values
+ `values_to` - Name of the new column; it will contain the data stored in the values of gathered columns

**Assess whether or not the data are tidy.**  
(1) each variable has its own column?  
(2) each observation has its own row? 
(3) each value has its own cell?  
  
**To fix this problem, we need to reshape the table to long format while keeping track of column names and values. We do this using `pivot_longer()`. Notice that the dimensions of the data frame change.**

```{r}
heartrate <- read_csv("/Users/lejla.becirevic/Desktop/BIS15W2024_lbecirevic/homework/data/heartrate.csv")
```

```{r}
heartrate %>% 
  pivot_longer(-patient, #patient will not move
               names_to = "drug", #make a new column called "drug"
               values_to="heartrate" #values moved to a new column called "heartrate"
               )
```

**Examples**
Import the file `relig_income.csv` and store it as a new object `relig_income`.  
```{r}
relig_income <- read_csv("/Users/lejla.becirevic/Desktop/BIS15W2024_lbecirevic/homework/data/relig_income.csv")
```

Use `pivot_longer()` to make the data tidy.  
```{r}
relig_income %>% 
  pivot_longer(-religion,
               names_to = "income",
               values_to = "total")
```

Some (but not all) of the column names are data. We also have NA's.
```{r}
billboard <- read_csv("/Users/lejla.becirevic/Desktop/BIS15W2024_lbecirevic/homework/data/billboard.csv")
```

Solution 1: specify a range of columns that you want to pivot.
```{r}
billboard2 <- 
  billboard %>% 
  pivot_longer(wk1:wk76, # a range of columns
               names_to = "week",
               values_to = "rank", 
               values_drop_na = TRUE #this will drop the NA's
               )
```

Solution 2: OR, specify columns that you want to stay fixed.
```{r}
billboard3 <- 
  billboard %>% 
  pivot_longer(-c(artist, track, date.entered), #specific columns to stay fixed
               names_to = "week",
               values_to = "rank",
               values_drop_na = TRUE
               )
```

Solution 3: identify columns by a prefix, remove the prefix and all NA's.
```{r}
billboard %>% 
   pivot_longer(
   cols = starts_with("wk"), #columns that start with "wk"
   names_to = "week",
   names_prefix = "wk",
   values_to = "rank",
   values_drop_na = TRUE)
```

Import `plant_data.csv` as a new object `plant_data`.  
```{r}
plant_data <- read_csv("/Users/lejla.becirevic/Desktop/BIS15W2024_lbecirevic/homework/data/plant_data.csv")
```

Use `pivot_longer()` to make the data tidy. Focus the data only on genotype, water_sched_prog, and greenhouse.  
```{r}
plant_data %>% 
  pivot_longer(-c(genotype, water_sched_prog, greenhouse),
               names_to = "day",
               values_to = "v2",
               values_drop_na = TRUE)
```

Recall that we use `pivot_longer()` when our column names actually represent variables. A classic example would be that the column names represent observations of a variable.
```{r}
datasets::USPersonalExpenditure
?USPersonalExpenditure
```

Here we add a new column of expenditure types, which are stored as rownames above, with `mutate()`. The `USPersonalExpenditures` data also needs to be converted to a data frame before we can use the tidyverse functions, because it comes as a matrix.
```{r}
expenditures <- USPersonalExpenditure %>% 
  as_tibble() %>% #this transforms the matrix into a data frame
  mutate(expenditure = rownames(USPersonalExpenditure))
```

Are these data tidy? Please use `pivot_longer()` to tidy the data.
```{r}
expenditures %>% 
  pivot_longer(-expenditure, #not moving expenditure 
               names_to = "year", #make new variable = year
               values_to = "bn_dollars") #moving values into bn_dollars
```

Restrict the data to medical and health expenditures only. Sort in ascending order.
```{r}
expenditures %>% 
  pivot_longer(-expenditure,
               names_to = "year",
               values_to = "bn_dollars") %>% 
  filter(expenditure=="Medical and Health") %>% 
  arrange(-bn_dollars)
```

```{r}
expenditures %>% 
  pivot_longer(-expenditure,
               names_to = "year",
               values_to = "bn_dollars") %>% 
  filter(expenditure=="Private Education") 
```

### names_sep 
**Helps pull these apart, but we still have "exp" and "rep" to deal with.**  
```{r}
qpcr_untidy <- read_csv("/Users/lejla.becirevic/Desktop/BIS15W2024_lbecirevic/homework/data/qpcr_untidy.csv")
```

```{r}
qpcr_untidy %>% 
  pivot_longer(
    exp1_rep1:exp3_rep3, #range
    names_to= c("experiment", "replicate"), #names of columns
    names_sep="_", #separate by _
    values_to="mRNA_expression")
```

### separate
In this new heart rate example, we have the sex of each patient included with their name. Are these data tidy? No, there is more than one value per cell in the patient column and the columns a, b, c, d once again represent values.
```{r}
heartrate2 <- read_csv("/Users/lejla.becirevic/Desktop/BIS15W2024_lbecirevic/homework/data/heartrate2.csv")
heartrate2
```

We need to start by separating the patient names from their sexes. `separate()` needs to know which column you want to split, the names of the new columns, and what to look for in terms of breaks in the data.
```{r}
heartrate2 %>% 
  separate(patient, into= c("patient", "sex"), sep = "_") #seperate patient from sex with the _
```

**Examples**
Re-examine `heartrate2`. Use `separate()` for the sexes, `pivot_longer()` to tidy, and `arrange()` to organize by patient and drug. Store this as a new object `heartrate3`.  
```{r}
heartrate3 <- heartrate2 %>% 
  separate(patient, into=c("patient", "sex"), sep="_") %>% 
  pivot_longer(-c(patient, sex),
               names_to = "drug",
               values_to = "heartrate")
heartrate3
```

### unite
**Opposite of separate(). Its syntax is straightforward. You only need to give a new column name and then list the columns to combine with a separation character.  Give it a try below by recombining patient and sex from `heartrate3`.**
```{r}
heartrate3 %>% 
  unite(patient_sex, "patient", "sex", sep=" ") #does the opposite, takes two variables and combines
```

### pivot_wider
The opposite of `pivot_longer()`. You use `pivot_wider()` when you have an observation scattered across multiple rows. In the example below, `cases` and `population` represent variable names not observations.  

Rules:  
+ `pivot_wider`(names_from, values_from)  
+ `names_from` - Values in the `names_from` column will become new column names  
+ `values_from` - Cell values will be taken from the `values_from` column  

```{r}
tb_data <- read_csv("/Users/lejla.becirevic/Desktop/BIS15W2024_lbecirevic/homework/data/tb_data.csv") #value column is not represented of single variable 
```

When using `pivot_wider()` we use `names_from` to identify the variables (new column names) and `values_from` to identify the values associated with the new columns.
```{r}
tb_data %>% 
  pivot_wider(names_from = "key", #the observations under key will become new columns
              values_from = "value") #the values under value will be moved to the new columns
```

**Examples**
Load the `gene_exp.csv` data as a new object `gene_exp`. Are these data tidy? Use `pivot_wider()` to tidy the data.
```{r}
gene_exp <- read_csv("/Users/lejla.becirevic/Desktop/BIS15W2024_lbecirevic/homework/data/gene_exp.csv")
```

```{r}
gene_exp %>% 
  pivot_wider(names_from = "type",
              values_from = "L4_values")
```

Load the beachbugs data and have a look.
```{r}
beachbugs <- read_csv("/Users/lejla.becirevic/Desktop/BIS15W2024_lbecirevic/homework/data/beachbugs_long.csv")
```

Use `pivot_wider` to transform the data into wide format.
```{r}
beachbugs_wide <- beachbugs %>% 
  pivot_wider(names_from = site,
              values_from = buglevels)
```

Now, use `pivot_longer` to transform them back to long!
```{r}
beachbugs_wide %>% 
  pivot_longer(-year,
               names_to = "site",
               values_to = "buglevels") %>% 
  arrange(desc(buglevels))
```

## LAB 10
### Grammar of Graphics
The ability to quickly produce and edit graphs and charts is a strength of R. These data visualizations are produced by the package `ggplot2` and it is a core part of the tidyverse. The syntax for using ggplot is specific and common to all of the plots. This is what Hadley Wickham calls a [Grammar of Graphics](http://vita.had.co.nz/papers/layered-grammar.pdf). The "gg" in `ggplot` stands for grammar of graphics.

### Philosophy
What makes a good chart? In my opinion a good chart is elegant in its simplicity. It provides a clean, clear visual of the data without being overwhelming to the reader. This can be hard to do and takes some careful thinking. Always keep in mind that the reader will almost never know the data as well as you do so you need to be mindful about presenting the facts.  

### Data Types
We first need to define some of the data types we will use to build plots.  

+ `discrete` quantitative data that only contains integers
+ `continuous` quantitative data that can take any numerical value
+ `categorical` qualitative data that can take on a limited number of values

### Basics
The syntax used by ggplot takes some practice to get used to, especially for customizing plots, but the basic elements are the same. It is helpful to think of plots as being built up in layers.  

In short, **plot= data + geom_ + aesthetics**.  

We start by calling the ggplot function, identifying the data, and specifying the axes. We then add the `geom` type to describe how we want our data represented. Each `geom_` works with specific types of data and R is capable of building plots of single variables, multiple variables, and even maps. Lastly, we add aesthetics.

**Example**
To make things easy, let's start with some built in data.
```{r}
names(iris)
```

```{r}
glimpse(iris)
```

To make a plot, we need to first specify the data and map the aesthetics. The aesthetics include how each variable in our data set will be used. In the example below, I am using the aes() function to identify the x and y variables in the plot.
```{r}
ggplot(data=iris, #specify the data
       mapping=aes(x=Species, y=Petal.Length)) #map the aesthetics
```

Notice that we have a nice background, labeled axes, and even a value range of our variables on the y-axis- but no plot. This is because we need to tell ggplot how we want our data represented. This is called the geometry or `geom()`. There are many types of `geom`, see the ggplot [cheatsheet](https://posit.co/resources/cheatsheets/).

Here we specify that we want a boxplot, indicated by `geom_boxplot()`.
```{r}
ggplot(data=iris, #specify the data
       mapping=aes(x=Species, y=Petal.Length))+ #map the aesthetics
  geom_boxplot() #add the plot type
```


Use the iris data to build a scatterplot that compares sepal length vs. sepal width. Use the cheat sheet for help to find the correct `geom_` for a scatterplot.
```{r}
names(iris)
```

```{r}
ggplot(data=iris,
       mapping = aes(x=Sepal.Length, y=Sepal.Width))+
  geom_point()
```


### Scatter Plots
Scatter plots are good at revealing relationships that are not readily visible in the raw data. For now, we will not add regression aka. "best of fit" lines or calculate any r^2^ values.  

In the case below, we are exploring whether or not there is a relationship between animal mass and home range. We are using the **log transformed values** because there is a large difference in mass and home range among the different species in the data.

Now that we have a general idea of the syntax, let's start by working with two common plots: 1) scatter plots and 2) bar plots.
```{r}
homerange <- read_csv("/Users/lejla.becirevic/Desktop/BIS15W2024_lbecirevic/homework/data/Tamburelloetal_HomeRangeDatabase.csv")
```

What is the structure of the `homerange` data? Does it have any NA's? Is it tidy? Do a quick exploratory analysis of your choice below.
```{r}
glimpse(homerange)
```

```{r}
names(homerange)
```

```{r}
ggplot(data=homerange, #specify the data
       mapping=aes(x=log10.mass, y=log10.hra))+ #map the aesthetics
  geom_point() #add the plot type
```

In big data sets with lots of overlapping values, over-plotting can be an issue. `geom_jitter()` is similar to `geom_point()` but it helps with over plotting by adding some random noise to the data and separating some of the individual points.
```{r}
ggplot(data=homerange, mapping=aes(x=log10.mass, y=log10.hra))+
  geom_jitter() #if you have really big data set (over plotting) => adds some random noise
```

To add a regression (best of fit) line, we just add another layer.
```{r}
ggplot(data=homerange, mapping=aes(x=log10.mass, y=log10.hra))+
  geom_point()+
  geom_smooth(method=lm, se=T) #add a regression line
```

**Examples**
What is the relationship between log10.hra and log10.preymass? What do you notice about how ggplot treats NA's?
```{r}
ggplot(data=homerange,
       mapping = aes(x=log10.hra, y=log10.preymass))+
  geom_point(na.rm=T)+
  geom_smooth(method=lm, se=T, na.rm = T)
```

### Bar Plot: geom_bar
The simplest type of bar plot counts the number of observations in a categorical variable. In this case, we want to know how many observations are present in the variable `trophic.guild`. Notice that we do not specify a y-axis because it is count by default.  

```{r}
names(homerange)
```

```{r}
homerange %>% 
  count(trophic.guild)
```

Also notice that we can use pipes! The `mapping=` function is implied by `aes` and so is often left out. 
```{r}
homerange %>% 
  ggplot(aes(trophic.guild)) +  #don't need to use x since we are counting one variable
  geom_bar() #good for counts and categorical data not continuous 
```

### Bar Plot: geom_col
Unlike `geom_bar()`, `geom_col()` allows us to specify an x-axis and a y-axis.
```{r}
homerange %>% 
  filter(family=="salmonidae") %>%
  select(common.name, log10.mass) %>% 
  ggplot(aes(y=common.name, x=log10.mass))+ #notice the switch in x and y => specified the y first
  geom_col()
```

`geom_bar()` with `stat="identity"`
`stat="identity"` allows us to map a variable to the y-axis so that we aren't restricted to counts.
```{r}
homerange %>% 
  filter(family=="salmonidae") %>% 
  ggplot(aes(x=common.name, y=log10.mass))+
  geom_bar(stat="identity") #only use when you want a count
```

**Examples**
Filter the `homerange` data to include `mammals` only.
```{r}
homerange %>% 
  filter(class=="mammalia")
```

Are there more herbivores or carnivores in mammals? Make a bar plot that shows their relative numbers.
```{r}
homerange %>% 
  select(class, trophic.guild) %>% 
  filter(class=="mammalia") %>% 
  ggplot(aes(trophic.guild))+
  geom_bar()
```

Make a bar plot that shows the masses of the top 10 smallest mammals.
```{r}
homerange %>% 
  filter(class=="mammalia") %>% 
  top_n(-10, log10.mass) %>% 
  ggplot(aes(x=common.name, y=log10.mass))+
  geom_col()+ #use when you have x and y axis
  coord_flip()
```

**Examples**
```{r}
penguins
```

In a previous lab, we asked how many penguins were measured on each island.
```{r}
penguins %>% 
  count(island)
```

Make this output more visual by adding a plot...
```{r}
penguins %>% 
  count(island) %>% 
  ggplot(aes(x=island, y=n))+
  geom_col()
```

```{r}
names(penguins)
```

What if we wanted a plot that showed the number of measured penguins for each species?
```{r}
penguins %>% 
  count(species) %>% 
  ggplot(aes(x=species, y=n))+
  geom_col()
```

```{r}
penguins %>% 
  ggplot(aes(species))+
  geom_bar()
```

How about average bill length by sex?
```{r}
penguins %>% 
  filter(sex!="NA") %>% 
  group_by(sex) %>% 
  summarise(avg_bill_lenth_mm=mean(bill_length_mm)) %>% 
  ggplot(aes(x=sex, y=avg_bill_lenth_mm))+
  geom_col()
```

### Box Plots
For the next series of examples, we will use the `homerange` data.  

```{r message=FALSE, warning=FALSE}
homerange <- read_csv("/Users/lejla.becirevic/Desktop/BIS15W2024_lbecirevic/homework/data/Tamburelloetal_HomeRangeDatabase.csv")
```

Boxplots help us visualize a range of values. So, on the x-axis we typically have something categorical and the y-axis is the range. In the case below, we are plotting `log10.mass` by taxonomic class in the `homerange` data. `geom_boxplot()` is the geom type for a standard box plot. The center line in each box represents the median, not the mean.

Let's look at the variable `log10.mass` grouped by taxonomic class.
```{r}
homerange %>% 
  group_by(class) %>% 
  summarize(min_log10.mass=min(log10.mass),
            max_log10.mass=max(log10.mass),
            median_log10.mass=median(log10.mass))
```

```{r}
homerange %>% 
  ggplot(aes(x = class, y = log10.mass)) +
  geom_boxplot()
```

**Examples**
There are more herbivores than carnivores in the homerange data, but how do their masses compare? Make a summary and boxplot that compares their masses. Use `log10.mass`.
```{r}
homerange %>% 
  select(trophic.guild, log10.mass) %>% 
  ggplot(aes(x=trophic.guild, y=log10.mass))+
  geom_boxplot()
```

Now use a boxplot to visualize the range of log10.mass by family of mammalian carnivore.
```{r}
homerange %>% 
  filter(trophic.guild=="carnivore" & class=="mammalia") %>% 
  ggplot(aes(x=family, y=log10.mass))+
  geom_boxplot()+
  coord_flip()
```






